White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-01-16 papers/wdmerger_II/paper.tex
Add some content to the numerical methods section in paper II.

% Software names
\newcommand{\boxlib}{\texttt{BoxLib}}
\newcommand{\castro}{\texttt{CASTRO}}
\newcommand{\wdmerger}{\texttt{wdmerger}}
\newcommand{\python}{\texttt{Python}}
\newcommand{\matplotlib}{\texttt{matplotlib}}
\newcommand{\yt}{\texttt{yt}}
\newcommand{\isoseven}{\texttt{iso7}}
\newcommand{\aproxthirteen}{\texttt{aprox13}}
\newcommand{\aproxnineteen}{\texttt{aprox19}}
\newcommand{\aproxtwentyone}{\texttt{aprox21}}

@@

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI}.
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator.

>>>>

% Software names
\newcommand{\boxlib}{\texttt{BoxLib}}
\newcommand{\castro}{\texttt{CASTRO}}
\newcommand{\microphysics}{\texttt{Microphysics}}
\newcommand{\wdmerger}{\texttt{wdmerger}}
\newcommand{\python}{\texttt{Python}}
\newcommand{\matplotlib}{\texttt{matplotlib}}
\newcommand{\yt}{\texttt{yt}}
\newcommand{\vode}{\texttt{VODE}}
\newcommand{\isoseven}{\texttt{iso7}}
\newcommand{\aproxthirteen}{\texttt{aprox13}}
\newcommand{\aproxnineteen}{\texttt{aprox19}}
\newcommand{\aproxtwentyone}{\texttt{aprox21}}

@@

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI}.
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator. There are three main areas of concern: first,
specifying what nuclides are in the network and what reaction rates link the
various nuclides; second, once the system of ODEs governing the evolution of
the nuclear species has been written down, how to append to them ODEs governing
the evolution of the thermodynamics, in particular the temperature; third,
coupling the nuclear reaction updates to the evolution of the hydrodynamical
system. We will deal with each of these in turn in the following subsections.

\subsection{Nuclear Network}
\label{sec:network}

White dwarfs are mainly composed of $\alpha$-chain particles, primarily ${}^4$He,
${}^{12}$C, ${}^{16}$O, ${}^{20}$Ne, and ${}^{24}$ Mg. Therefore the core of
any network appropriate for modeling nuclear burning in white dwarfs will be
these alpha chain nuclides, with the idea being that links up the $\alpha$-chain
will eventually get us to ${}^{56}$Ni, the nuclide responsible for the
energy output of Type Ia supernovae. In this paper we consider four networks
to do this, presented in order of increasing complexity. The most simple is
\isoseven\ \citep{timmes:2000}, which includes all of the aforementioned isotopes and
${}^{28}$Si (see also \citet{hix:1998}). ${}^{28}$Si effectively measures the
equilibrium state of silicon-group elements, and ${}^{56}$Ni effectively measures
the equilibrium state of iron-group elements, with the link between them governed
by the effective loss or gain of seven $\alpha$-particles. This type of network
was used by \citet{rosswog:2009} for their collision calculations in SPH.

Next is \aproxthirteen\ \citep{timmes:1999,timmes:2000}. This includes
all of the isotopes of \isoseven, and all of the $alpha$-chain particles between
silicon and nickel (${}^{32}$S, ${}^{36}$Ar, ${}^{40}$Ca, ${}^{44}$Ti, ${}^{48}$Cr,
and ${}^{52}$Fe). This network was used by \citet{hawley:2012} and \citet{raskin:2010}.
\citet{loren-aguilar:2010} and \citet{garcia-senz:2013} used a very similar network
that included additionally ${}^{60}$Zn. The \aproxnineteen\ network \citep{timmes:1999}
builds on \aproxthirteen\ by including isotopes for hydrogen burning and explicit
tracking of photodisintegration into ${}^{54}$Fe. This network was used by
\citet{kushnir:2013}, \citet{kushnir:2014}, \citet{rosswog:2009} for their
calculations with FLASH, and \citet{papish:2015}. Finally we will also
consider \aproxtwentyone, which includes all of the above plus ${}^{56}$Cr
and ${}^{56}$Fe and related reaction links. The primary virtue of using
the latter two networks is that they allow us to track changes away from
$Y_e = 0.5$.

All four of these networks have been ported into a form that is consistent
with the \boxlib\ codes, in the freely available \microphysics\ code
repository\footnote{\microphysics\ can be obtained at \url{https://github.com/BoxLib-Codes/Microphysics}.},
a collection of microphysical routines that are designed to be used in our
hydrodynamics codes. These can be easily swapped at compile time by using the 
appropriate makefile variable.

\subsection{Nuclear Burning}
\label{sec:burner}

Given a set of nuclides and the reaction links between them, we now consider
how a burning step is performed in our software. The integration of the stiff,
implicit system is performed by \vode\ \citep{vode}, a copy of which is
provided with \castro. To \vode\ we provide an integration state containing
the molar abundances $Y_{n} = X_{n} / A_{n}$, where $X$ is the mass fraction
of the abundance in the zone and $A$ is the number of nucleons. The integration
of
\begin{equation}
  \frac{d\bm{Y}}{dt} = f(\mathbf{Y})
\end{equation}
is determined using a right-hand-side provided by the nuclear burning network.
When it comes to the thermodynamical evolution, the internal energy $e$ of the zone
will change when the nuclear abundances evolve. During the integration we track
the energy release from the system using
\begin{equation}
  \frac{\partial e}{\partial t} = N_A \sum_{n} \frac{\partial Y_{n}}{\partial t} m_{n} c^2,
\end{equation}
where $c$ is the speed of light and $m_n$ is the mass of each nuclide.

In a hydrostatic burn, we keep $\rho$ and $T$ fixed throughout, and use
the energy released at the end to compute a final temperature that is
thermodynamically consistent with the new internal energy. By contrast,
in a self-heating burn, we allow the temperature to evolve in response
to the burning (see \citet{maestro3}):
\begin{equation}
  \frac{dT}{dt} = \frac{1}{c_V}\left(\frac{\partial e}{\partial t} - \sum_n \frac{\partial e}{\partial X_n}\frac{\partial X_n}{\partial t}\right)
\end{equation}
Here $c_V$ is the specific heat at constant volume, and $(\partial e/\partial X_n)$
is the derivative of the internal energy with respect to the mass fraction of a given
species (with other thermodynamic variables held constant). Both quantities are provided by the
equation of state. During this burn, we can keep these thermodynamic
derivatives constant using their value at the beginning of the run, or at each step we
can choose to re-evaluate the equation of state using the latest value of $(rho, T)$.
The latter is more expensive but also more accurate, and we use it in this paper.
A third option presented by \citet{raskin:2010} is a so-called ``hybrid'' mode.
In this mode, by default we do a hydrostatic burn. If that burn fails, or if the net
energy change is negative, we do the burn again in self-heating mode. All three options
are implemented in our burner software.

\subsection{Coupling to the Hydrodynamics}
\label{sec:hydrocoupling}

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here.
(A related approach presented in \citet{kushnir:2013} is to limit the magnitude
of the energy release in any given integration by scaling all the reaction rates.)
At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum burning
timestep. In a given zone, the burning timestep is defined as:
\begin{equation}
  \Delta t_{\text{burning}} = f_b\, \frac{e}{\dot{e}}.
\end{equation}
Here $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning. We estimate this using the energy change in the last burning
update, divided by the length of the burning timestep. The factor $f_b$ then
determines by what fraction we would like to allow the internal energy to change
in the current timestep (assuming that $\dot{e}$ does not change quickly from
timestep to timestep, which is not always the case). By making $f_b$ smaller,
we can control the magnitude of the decoupling between the reactions and the hydro.
A typical choice in the literature is $f_b \sim 0.3$. This timestep limiting has
now been implemented in \castro\ and the factor $f_b$ can be set at runtime.
